Compilation and Backend-Independent Vectorization for Multi-Party Computation

### 2022-05-02 overleaf/model.tex
Final edits. Will freeze TR.

%\ana{We need an intro to section here.}
\ana{Terminology: "operation" or "instruction" or "gate"?}

@@

\begin{enumerate}
\item All statements in the program execute using the same protocol (sharing). That is, there is no share conversion.
%\ishaq{This is not an assumption, this is our setting. We work in the single protocol world.}
\item There are two tiers of MPC instructions, local and remote. A local instruction (essentially just ADD) has cost $\beta$
and a remote instruction (e.g., MUX, MUL, SHL, etc.) has cost $\alpha$, where $\alpha >> \beta$. We assume that all remote
instructions have the same cost.
%\ishaq{Don't we need to distinguish between local (cheap) vs. interactive (expensive) instructions? I don't see how we could assume this. -- Past suggestion from Vassilis: Ideally we should assume cost to be a convex function.}\ana{After discussion 1/12: replace single costs 1 with two levels of costs: $\alpha$ for non-local operations, e.g., MUL, and $\beta$ for local ops, e.g., ADD.}
\item  In MPC frameworks, executing $n$ operations ``at once'' in a single SIMD operation costs a lot less than executing those $n$ operations one by one.
Following Amdahl's law~\ana{citation!}, we write $\alpha = \frac{1}{s}p\alpha + (1-p)\alpha$, where $p$ is the fraction of execution time that benefits from amortization and $(1-p)$
is the fraction that does not, and $s$ is the available resource. Thus, $n\alpha = \frac{n}{s}p\alpha + n(1-p)\alpha$.
For the purpose of the model we assume that $s$ is large enough and the term $\frac{n}{s}p\alpha$ amounts to a \emph{fixed cost} incurred regardless of
whether $n$ is $10,000$ or just $1$. (This models the cost of preparing and sending a packet from party A to party B for example.) Therefore, amortized execution
of $n$ operations is $f(n) = \alpha_\mathit{fix} + n\alpha_{var}$ in contrast to unamortized execution $g(n) = n\alpha_\mathit{fix} + n\alpha_{var}$ and, of course.
Clearly $\alpha_\mathit{fix} << n\alpha_\mathit{fix}$ and since fixed cost dominates variable cost (particularly for remote operations), we have $f(n) << g(n)$.
%We assume infinite parallel capacity---i.e., a single MPC-instruction costs as much as $N$ amortized instructions, namely $\alpha$ or $\beta$.
%This is a standard assumption in Cryptographic Parallel RAM. ABY presents empirical support for this assumption~\ana{Add citations. PRAM, ABY.}
%\ishaq{Comment from Vassilis: We assume there are infinite parallel capacities, this is the standard assumption in PRAM (Cryptographic Parallel RAM).}\ana{PRAM assumption stays. We have to replace the unit cost of 1 with a suitable amortization function $f(n)$, which I don't think changes much in the cost modeling and analysis.}
\item MPC instructions scheduled in parallel benefit from amortization \emph{only if} they are the same instruction. Given our previous assumption,
2 MUL instructions can be amortized in a single SIMD instruction that costs $\alpha_\mathit{fix} + 2\alpha_\mathit{var}$, however a MUL and a MUX instruction
still cost $2\alpha_\mathit{fix} + 2\alpha_\mathit{var}$ even when scheduled ``in parallel''.\footnote{This is not strictly true, but assuming it, e.g. as in 
\cite{Ishaq2019, Demmler2015ABYA, Mohassel2018}, helps simplify the problem.}
%MPC instructions scheduled in parallel benefit from amortization \emph{only if} they are the same instruction. Given our previous assumption,
%2 MUL instructions scheduled in parallel benefit from amortization and cost $\alpha$, however a MUL and a MUX instructions scheduled
%in parallel still cost $2\alpha$.
%\ishaq{We need to reword this assumption to something like "$K$ parallel MUL costs much less than $K$ (because they get amortized), but any mix of $K$  MUL and MUX still cost roughly $K$. Specifically, we should take away the low constants (2 and 1) because we know for low constants this is not true.} \ana{Again, core assumption that MUL and MUX don't benefit from amortization stays. We have to replace constant costs with functions, as in (3).}
\end{enumerate}

>>>>

%\ana{We need an intro to section here.}
%\ana{Terminology: "operation" or "instruction" or "gate"?}

@@

\begin{enumerate}
\item All statements in the program execute using the same protocol (sharing). That is, there is no share conversion.
%\ishaq{This is not an assumption, this is our setting. We work in the single protocol world.}
\item There are two tiers of MPC instructions, local and remote. A local instruction (e.g., ADD in Arithmetic, XOR in Boolean) 
has cost $\beta$ and a remote instruction (e.g., MUX, MUL, SHL, etc.) has cost $\alpha$, where $\alpha >> \beta$. We assume that all remote
instructions have the same cost.
%\ishaq{Don't we need to distinguish between local (cheap) vs. interactive (expensive) instructions? I don't see how we could assume this. -- Past suggestion from Vassilis: Ideally we should assume cost to be a convex function.}\ana{After discussion 1/12: replace single costs 1 with two levels of costs: $\alpha$ for non-local operations, e.g., MUL, and $\beta$ for local ops, e.g., ADD.}
\item  In MPC frameworks, executing $n$ operations ``at once'' in a single SIMD operation costs a lot less than executing those $n$ operations one by one.
Following Amdahl's law, we write $\alpha = \frac{1}{s}p\alpha + (1-p)\alpha$, where $p$ is the fraction of execution time that benefits from amortization and $(1-p)$
is the fraction that does not, and $s$ is the available resource. Thus, $n\alpha = \frac{n}{s}p\alpha + n(1-p)\alpha$.
For the purpose of the model we assume that $s$ is large enough and the term $\frac{n}{s}p\alpha$ amounts to a \emph{fixed cost} incurred regardless of
whether $n$ is $10,000$ or just $1$. (This models the cost of preparing and sending a packet from party A to party B for example.) Therefore, amortized execution
of $n$ operations is $f(n) = \alpha_\mathit{fix} + n\alpha_{var}$ in contrast to unamortized execution $g(n) = n\alpha_\mathit{fix} + n\alpha_{var}$.
We have $\alpha_\mathit{fix} << n\alpha_\mathit{fix}$ and since fixed cost dominates variable cost (particularly for remote operations), we have $f(n) << g(n)$.
%We assume infinite parallel capacity---i.e., a single MPC-instruction costs as much as $N$ amortized instructions, namely $\alpha$ or $\beta$.
%This is a standard assumption in Cryptographic Parallel RAM. ABY presents empirical support for this assumption~\ana{Add citations. PRAM, ABY.}
%\ishaq{Comment from Vassilis: We assume there are infinite parallel capacities, this is the standard assumption in PRAM (Cryptographic Parallel RAM).}\ana{PRAM assumption stays. We have to replace the unit cost of 1 with a suitable amortization function $f(n)$, which I don't think changes much in the cost modeling and analysis.}
\item MPC instructions scheduled in parallel benefit from amortization \emph{only if} they are the same instruction. Given our previous assumption,
2 MUL instructions can be amortized in a single SIMD instruction that costs $\alpha_\mathit{fix} + 2\alpha_\mathit{var}$, however a MUL and a MUX instruction
still cost $2\alpha_\mathit{fix} + 2\alpha_\mathit{var}$ even when scheduled ``in parallel''.\footnote{This is not strictly true, but assuming it, e.g. as in 
\cite{Ishaq:2019, NDSS:DemSchZoh15, Mohassel:2018}, helps simplify the problem.}
%MPC instructions scheduled in parallel benefit from amortization \emph{only if} they are the same instruction. Given our previous assumption,
%2 MUL instructions scheduled in parallel benefit from amortization and cost $\alpha$, however a MUL and a MUX instructions scheduled
%in parallel still cost $2\alpha$.
%\ishaq{We need to reword this assumption to something like "$K$ parallel MUL costs much less than $K$ (because they get amortized), but any mix of $K$  MUL and MUX still cost roughly $K$. Specifically, we should take away the low constants (2 and 1) because we know for low constants this is not true.} \ana{Again, core assumption that MUL and MUX don't benefit from amortization stays. We have to replace constant costs with functions, as in (3).}
\end{enumerate}

### 2022-05-02 overleaf/paper_SIMD.tex
Final edits. Will freeze TR.

\ana{We need an abstract.}

@@

\ana{Need a conclusion here!}

>>>>

Research on MPC programming technology largely falls at the two ends of the classical compiler: (1) work on front-end language design (e.g., Wysteria, Viaduct) and (2) work on back-end protocol implementation (e.g., ABY, MOTION).

In this work, we formalize the MPC Source intermediate language and advance the idea of what we call backend-independent optimizations, in a close analogy to machine-independent optimizations in the classical compiler.  We present a compiler framework that takes a Python-like routine and produces MOTION code. We focus on a specific backend-independent optimization: novel SIMD-vectorization on MPC Source, which we show leads to significant speedup in circuit generation time and running time, as well as significant reduction in communication and number of gates over the corresponding iterative schedule.
%Our results emphasize the importance of backend-independent optimizations --- vectorization (this work) and protocol mixing (previous work) are two optimizations available at the level of MPC Source. We envision future work on backend-independent optimization: new MPC optimizations and combinations of optimizations in the vein of the classical compiler.

@@

We presented a formalization of the MPC Source intermediate language followed by a specific back-end optimization at the level of MPC Source: novel SIMD-vectorization. We demonstrated that vectorization has significant impact on performance. 
We are excited about the opportunities for future work --- integration with protocol mixing, divide-and-conquer reasoning and parallelization, as well as inter-procedural context-sensitive analysis at the level of MPC Source will improve MPC programmability and efficiency.

### 2022-05-02 overleaf/related.tex
Final edits. Will freeze TR.

HyCC~\cite{CCS:BDKKS18} is a compiler from C Source into ABY circuits. It does source-to-source compilation with the key goal to decompose the 
program into modules and then assign protocols to modules. In contrast, we focus on MPC Source-level optimizations, specifically vectorization, although we envision 
future optimizations as well. We formalize MPC Source and reasoning about transformations, which we conjecture is more tractable than reasoning over
the higher-level AST. On the other hand, HyCC does inter-procedural optimizations, while our analysis is intra-procedural. We will explore context-sensitive inter-procedural 
analysis over MPC Source in future work. HyCC, similarly to Buscher~\cite{Buscher:2018} uses an of-the-shelf source-to-source polyhedral compiler~\footnote{We believe HyCC uses 
Par4All~\ana{add citation!}, however, it is not appear to be included with the publicly available distribution of HyCC.}
to perform vectorization at the level of source code. The disadvantage of using an of-the-shelf source-to-source compiler is that it solves a more general 
problem than what MPC presents and may forgo opportunities for optimization --- concretely, it is well-known that vectorization and polyhedral compilation 
do not work well with conditionals~\cite{Benabderrahmane:2010,Karrenberg:2015}. In contrast, we consider vectorization at the level of MPC Source which 
linearizes conditionals; we are able to handle programs with interleaved if-statements and for-loops and achieve significant speedup.

>>>>

HyCC~\cite{CCS:BDKKS18} is a compiler from C Source into ABY circuits. It does source-to-source compilation with the key goal to decompose the 
program into modules and then assign protocols to modules. In contrast, we focus on MPC Source-level optimizations, specifically vectorization, although we envision 
future optimizations as well. We formalize MPC Source and reasoning about transformations, which we conjecture is more tractable than reasoning over
the higher-level AST. On the other hand, HyCC does inter-procedural optimizations, while our analysis is intra-procedural. We will explore context-sensitive inter-procedural 
analysis over MPC Source in future work. HyCC, similarly to Buscher~\cite{Buscher:2018} uses an of-the-shelf source-to-source polyhedral compiler~\footnote{We believe HyCC uses 
Par4All (\url{https://github.com/Par4All/par4all}), however, does not appear to be included with the publicly available distribution of HyCC.}
to perform vectorization at the level of source code. The disadvantage of using an of-the-shelf source-to-source compiler is that it solves a more general 
problem than what MPC presents and may forgo opportunities for optimization --- concretely, it is well-known that vectorization and polyhedral compilation 
do not work well with conditionals~\cite{Benabderrahmane:2010,Karrenberg:2015}. In contrast, we consider vectorization at the level of MPC Source which 
linearizes conditionals; we are able to handle programs with interleaved if-statements and for-loops and achieve significant speedup.

### 2022-05-02 overleaf/results.tex
Final edits. Will freeze TR.

A detailed summary of the effects of vectorization on various benchmarks is presented in \cref{table:metrics}. We show circuit evaluation times in \cref{fig:graph_all_eval_time}. In terms of amenability to vectorization, we divide benchmarks into 3 categories: 1) {\it High:} these include convex hull, cryptonets max pooling, minimal points and private set intersection. These benchmarks are highly parallelizable and see 25x to 70x speedup in BMR, and 30x to 55x in GMW protocol. 2) {\it Medium:} these include biometric matching, DB Variance, histogram, inner product, k-means iteration and MNIST ReLU. These benchmarks have non-parallelizable phases e.g. the summing phase of inner product and biometric matching. Still, most computation is parallelizable and it results in speedup from 5x to 25x in BMR, and 2x to 25x in GMW protocol. 3) {\it Low:} these include the Database Join and the regular expression benchmarks (count 102, count 10, longest 102 and max distance between symbols). There is less parallelizable computation in these programs, thus the speedup is lower. We see a speedup from 1.1x to 2x in BMR. In GMW, DB Join, Count 102 and Count 10s see speedup from 1.1x to 1.3x. However, longest102 and max distance between symbols suffer a slowdown of 0.5x. There is opportunity for vectorization in these benchmarks according to our analytical model, particularly, a large EQ operation is vectorized, though a large portion of the loop cannot be vectorized. We observed that transformation to vectorized code increased multiplicative depth and, the negative effect of increased depth is more noticeable in a round-based protocol like GMW. The cause of the increase is not clear --- we conjecture that MOTION performs optimizations over the non-vectorized loop body that decreases depth; also, EQ is relatively inexpensive in Boolean GMW and BMR compared to ADD and MUL, which also de-emphasizes the benefit of vectorization. We propose a simple heuristic: if the transformation increases circuit depth beyond some threshold (e.g. more than 10\% of the original circuit), we reject the transformation. %Nevertheless, we show these graphs here for the sake of completeness and to highlight that vectorization is not always reduce run time. 
Note that in some settings it may still be desirable to vectorize e.g. in data constrained environments. As shown in \cref{fig:graph_comm_size}, vectorization results in reduced communication (fewer bits are transferred).

@@

%We conclude the . 
Finally, we compared our automatically generated routine for Inner Product with the manually SIMD-ified MOTION-native routine in the distribution. 
We were surprised that we were an order of magnitude slower in Boolean GMW as our circuit ran a significantly larger 
number of communication rounds. Upon investigation, it turned out that the vectorized multiplications were essentially the same, 
however, our addition loop incurred significant cost (ADD is non-local and expensive in Boolean GMW). The MOTION-native loop ran 
{\sf result += mult\_unsimdified[i];} while our loop generated and ran {\sf result[i] = result[i-1] + mult\_unsimdified[i];} (recall that the array computation
is an artifact of our vectorization). We rewrote the accumulation (manually, for testing purposes) and that led to the comparable speedup!

Therefore, MOTION's compiler performs analysis that informs circuit generation and the example illustrates the power of the analysis. In the above example, 
MOTION does the standard divide-and-conquer accumulation. The poor performance of our loop is not a limitation of our approach, but of our current 
implementation. Recall that the next phase of Vectorization, in~\secref{sec:vectorization}, which we have not implemented yet, gets rid of redundant 
dimensions and generates {\sf result = result\_prev + mult\_unsimdified[i]}. This could have benefited from MOTION's divide-and-conquer optimization
but it might lead to confusion of the static analysis as analysis on the AST is difficult (unfortunately, we ran out of time and did not experiment further).
%Interestingly, the non-vectorized version, which is the literal translation of inner product does not appear to benefit from any optimizations.

We conjecture that MPC source, as a more straight-forward representation will not only enable detection of general associative loops, but also allow 
for program synthesis to increase opportunities for divide-and-conquer parallelization~\cite{Farzan:2021}; we leave this as future work.

The most closely related work, HyCC~\cite{CCS:BDKKS18}, is a mixing compiler that focuses on mixed protocols, while we run vectorization within a single protocol. 
The paper only provides data for Boolean and Yao for a version of Biometric matching on $N=1000$; it does provide a lot of data on mixed protocol executions.
Again, we estimate we are about an order of magnitude slower. This is likely due to the same issue as Inner Product --- the computation of min can be optimized.

>>>>

A detailed summary of the effects of vectorization on various benchmarks is presented in \cref{table:metrics}. We show circuit evaluation times in \cref{fig:graph_all_eval_time}. In terms of amenability to vectorization, we divide benchmarks into 3 categories: 1) {\it High:} these include convex hull, cryptonets max pooling, minimal points and private set intersection. These benchmarks are highly parallelizable and see 25x to 70x speedup in BMR, and 30x to 55x in GMW protocol. 2) {\it Medium:} these include biometric matching, DB Variance, histogram, inner product, k-means iteration and MNIST ReLU. These benchmarks have non-parallelizable phases e.g. the summing phase of inner product and biometric matching. Still, most computation is parallelizable and it results in speedup from 5x to 25x in BMR, and 2x to 25x in GMW protocol. 3) {\it Low:} these include the Database Join and the regular expression benchmarks (count 102, count 10, longest 102 and max distance between symbols). There is less parallelizable computation in these programs, thus the speedup is lower. We see a speedup from 1.1x to 2x in BMR. In GMW, DB Join, Count 102 and Count 10s see speedup from 1.1x to 1.3x. However, longest102 and max distance between symbols suffer a slowdown of 0.5x. There is opportunity for vectorization in these benchmarks according to our analytical model, particularly, a there is a large EQ that is vectorized, although a large portion of the loop cannot be vectorized. We observed that transformation to vectorized code increased multiplicative depth and, the negative effect of increased depth is more noticeable in a round-based protocol like GMW. The cause of the increase is not clear --- we conjecture that MOTION performs optimizations over the non-vectorized loop body that decreases depth; also, EQ is relatively inexpensive in Boolean GMW and BMR compared to ADD and MUL, which also de-emphasizes the benefit of vectorization. We propose a simple heuristic although we do leave all the benchmarks in the table: if the transformation increases circuit depth beyond some threshold (e.g. more than 10\% of the original circuit), we can reject the transformation. %Nevertheless, we show these graphs here for the sake of completeness and to highlight that vectorization is not always reduce run time. 
Note that in some settings it may still be desirable to vectorize e.g. in data constrained environments. As shown in \cref{fig:graph_comm_size}, vectorization results in reduced communication (fewer bits are transferred).

@@

%We conclude the . 
Finally, we compared our automatically generated routine for Inner Product with the manually SIMD-ified MOTION-native routine in the distribution. 
We were surprised that we were an order of magnitude slower in Boolean GMW as our circuit ran a significantly larger 
number of communication rounds. Upon investigation, it turned out that the vectorized multiplications were essentially the same, 
however, our addition loop incurred significant cost (ADD is non-local and expensive in Boolean GMW). The MOTION-native loop ran 
{\sf result += mult\_unsimdified[i];} while our loop generated and ran {\sf result[i] = result[i-1] + mult\_unsimdified[i];} recall that the scalar expansion
is an artifact of our vectorization. We rewrote the accumulation (manually, for testing purposes) and that led to the comparable speedup!

MOTION's compiler performs analysis that informs circuit generation and the example illustrates the power of the analysis. In the above example, 
MOTION does the standard divide-and-conquer accumulation. The poor performance of our loop is not a limitation of our approach, but it could be a 
limitation of our current implementation. Recall that the next phase of Vectorization, in~\secref{sec:vectorization}, which we have not implemented yet, gets rid of redundant 
dimensions and generates {\sf if (i!=0) { result\_pref = result; } result = result\_prev + mult\_unsimdified[i]}. And while it is unrealistic to expect that MOTION's static analysis 
will detect the associative accumulation in the scalar expansion code, it is realistic to expect that it will in the above only slightly more complex code. 
It still might lead to confusion in the static analysis as analysis on the AST is difficult (unfortunately, we ran out of time and did not experiment further).
%Interestingly, the non-vectorized version, which is the literal translation of inner product does not appear to benefit from any optimizations.

We conjecture that MPC Source, a straight-forward representation, will not only enable detection of general associative loops, but also allow 
for program synthesis to increase opportunities for divide-and-conquer parallelization~\cite{Farzan:2021}; we leave this as future work.

The most closely related work, HyCC~\cite{CCS:BDKKS18}, is a mixing compiler that focuses on mixed protocols, while we run vectorization within a single protocol. 
The paper only provides data for Boolean and Yao for a version of Biometric matching on $N=1000$; it does provide a lot of data on mixed protocol executions.
Again, we estimate we are about an order of magnitude slower, which is likely due to the same issue as Inner Product --- the computation of $\mathit{min}$ can be optimized.
